---
title: "Vietnam FMD DAPC analysis"
format:
  html: 
    df-print: kable
    code-fold: show
    code-summary: "Hide code"
    code-overflow: wrap
    html-math-method: katex
    smooth-scroll: true
editor: source
editor_options: 
  chunk_output_type: console
---


## Overview
Script below applies a DAPC and spatial modeling techniques to estimate virus diversity at unsampled locations.  The goal to map FMDv diversity across a grid and to estimate diversity within provinces where virus was not collected.
  

## Load Libraries
```{r packages, warning=FALSE, message=FALSE}
library(here) #directory management
library(tidyverse) #general wrangling
library(ape) #phylodata manipulation
library(phangorn) #phylodata manipulation
library(msa) #phylodata manipulation
library(adegenet) #DAPC package

# OSF
library(osfr) # data management and archiving

  
##for spatial analyses and disaggregation
library(sp) #spatial data manipulation
library(spdep) #spatial data manipulation
library(terra) #spatial data manipulation
library(sf) #spatial data manipulation
library(geosphere) #distance measurements
library(vegan) #Mantel tests and Redundancy Analysis
library(ggforce) #help with pie chart plots
library(ggpubr) #help with pie chart plots
library(viridis) #color pallete
library(ggtree) #plot trees
library(patchwork) #combine plots
library(cowplot) #pull legend from ggplot
library(geodata) #import spatial covaraites
library(INLA) #model inference
library(Matrix) #diagnostics to check matrices
```

## Read Alignments and Data  
```{r data, warning=FALSE, message=FALSE}
#sequence metadata: sequence accession names with associated location/date/host metadata
load(here("data/prep-for-seqs.Rda")) #data object of interest: seq_meta
rm(seqs.g23, seqs.csv1, genbank_return1, genbank_return2, seq_meta_orig, prov_reg, seq_meta_reg)

#clean trimmed VP1 alignments
#might need to revisit outlier removal process eventually, but these should work for now
alignment.SEA<-read.dna(here("data/vp1_O-SEA_trimmed.fasta"),
                        format="fasta", as.matrix=TRUE)

alignment.Cath<-read.dna(here("data/vp1_O-Cath_trimmed.fasta"),
                         format="fasta", as.matrix=TRUE)

alignment.MESA<-read.dna(here("data/vp1_O-MESA_trimmed.fasta"),
                         format="fasta", as.matrix=TRUE)

alignment.A<-read.dna(here("data/vp1_A_trimmed.fasta"),
                      format="fasta", as.matrix=TRUE)

#Vietnam shapefile: country shapefile with provinces in equal area projection
load(here("data/vn-spat.Rda")) #data object of interest: vn_provinces_eq
rm(vn_provinces, vn_spat1, vn_prov_reg1)

#equal area projection for reference
equal_area_proj<-"+proj=aea +lat_1=20 +lat_2=40 +lat_0=30 +lon_0=105 +x_0=0 +y_0=0 +datum=WGS84 +units=km +no_defs"
```


## Example DAPC
Example to run DAPC on `alignment.SEA`.  There's a *run_dapc_alignment()* function below that captures the steps in this chunk. The code here was the stepwise process used to piece it together.
```{r eval=FALSE}
# get metadata for target alignment
meta <- seq_meta %>%
  filter(accession %in% labels(alignment.SEA)) %>%
  distinct(accession, .keep_all = TRUE)

aln <- alignment.SEA[meta$accession, ]

# province as factor, count samples
meta <- meta %>%
  mutate(province = factor(location)) %>% # location to factor
  group_by(province) %>% # group
  mutate(n_in_group = n()) %>% # sample count by province
  ungroup()

kept <- meta$n_in_group >= 1 # keep provinces with samples, can try 2 later
meta <- meta[kept, ] # filter/subset to provinces with samples
aln <- aln[kept, ] # filter/subset to provinces with samples
grp <- droplevels(meta$province) # filter/subset to provinces with samples

length(grp) # samples
length(unique(grp)) # unique provinces

# convert to "genind" object used by adegenet package
gi <- adegenet::DNAbin2genind(aln, polyThres = 0)

# allele matrix
# tab() expands genotypes to numeric dummy variables
# NA.method="mean" imputes missing as the mean allele frequency
allele_mat <- tab(gi, NA.method = "mean")  # n x p numeric

# scale alleles (so each locus has variance ~1)
allele_mat <- scale(allele_mat, center = TRUE, scale = TRUE)

## Run DAPC
set.seed(1976)

# how many PCs to use
n_pca_max <- nrow(allele_mat) - length(levels(grp))

dapc_initial_fit <- dapc(
  allele_mat, # allele matrix
  grp, # province group
  n.pca = n_pca_max,# max PC to use
  n.da  = 5
)

# proportion provinces correctly estimated using PCs
summary(dapc_fit)$assign.prop 

# optimum number of PCs to use
optim_pcs <- optim.a.score(dapc_initial_fit)
optim_pcs$best

dapc_final_fit <- dapc(
  allele_mat, # allele matrix
  grp, # province group
  n.pca = optim_pcs$best,# max PC to use
  n.da  = 5
)

# proportion correct
summary(dapc_final_fit)$assign.prop

# save main results

# PCs used by DAPC (each sample)
keep_PCs <- dapc_final_fit$tab
names(keep_PCs) <- paste0("PC_", 1:dim(keep_PCs)[2])
keep_PCs$virus <- rownames(keep_PCs)

# Linear Discriminates (each sample)
keep_LDs <- as.data.frame(dapc_final_fit$ind.coord)
names(keep_LDs) <- paste0("LD_", 1:dim(keep_LDs)[2])
keep_LDs$virus <- rownames(keep_LDs) 

# Linear Discriminates (for groups, i.e., provinces)
keep_grpLDs <- as.data.frame(dapc_final_fit$grp.coord)
names(keep_grpLDs) <- paste0("LD_", 1:dim(keep_grpLDs)[2])
keep_grpLDs$province <- rownames(keep_grpLDs) 
```


## Run DAPC Function  
The *run_dapc_alignment()* function performs the above steps.  

   
Saving results to project folder on OSF
```{r}
personal_osf_api <- yaml::read_yaml(here("local", "secrets.yaml")) # PAT credentials
osf_auth(personal_osf_api$osf_token) # register PAT for write/admin privileges

osf_project_data <- osf_retrieve_node("https://osf.io/ns9t3/")

osf_project_data$name # OSF directory for "FMDV Vietnam" project
```

### Run for SEA
```{r eval=FALSE}
dapc_SEA <- run_dapc_alignment(alignment = alignment.SEA,
                               seq_meta,
                               location_col = "location",
                               accession_col = "accession",
                               min_samples_per_group = 1,
                               n_da = 5,
                               seed = 1976)

names(dapc_SEA)

# save local copy
saveRDS(dapc_SEA, here("local/dapc_2026-01-15/dapc_SEA_2026-01-15.rds"))

# push copy to OSF
osf_upload(osf_project_data, 
           conflicts = "overwrite",
           path = here("local/dapc_2026-01-15/dapc_SEA_2026-01-15.rds"))
```

### Run fo CATH
```{r eval=FALSE}
dapc_CATH <- run_dapc_alignment(alignment = alignment.Cath,
                               seq_meta,
                               location_col = "location",
                               accession_col = "accession",
                               min_samples_per_group = 1,
                               n_da = 5,
                               seed = 1976)

names(dapc_CATH)

# save local copy
saveRDS(dapc_CATH, here("local/dapc_2026-01-15/dapc_CATH_2026-01-15.rds"))

# push copy to OSF
osf_upload(osf_project_data, 
           conflicts = "overwrite",
           path = here("local/dapc_2026-01-15/dapc_CATH_2026-01-15.rds"))
```


### Run fo Serotype A
```{r eval=FALSE}
dapc_A <- run_dapc_alignment(alignment = alignment.A,
                             seq_meta,
                             location_col = "location",
                             accession_col = "accession",
                             min_samples_per_group = 1,
                             n_da = 5,
                             seed = 1976)

names(dapc_A)

# save local copy
saveRDS(dapc_A, here("local/dapc_2026-01-15/dapc_A_2026-01-15.rds"))

# push copy to OSF
osf_upload(osf_project_data, 
           conflicts = "overwrite",
           path = here("local/dapc_2026-01-15/dapc_A_2026-01-15.rds"))
```




## Spatial analyses

### Download DAPC Results

Download SEA DAPC results from OSF (if needed)
```{r eval=FALSE}
osf_id <- osf_project_data %>% # OSF project
  osf_ls_files() %>% # items in OSF project
  filter(name == "dapc_SEA_2026-01-15.rds") # file to retrieve

osf_download(osf_id,
             path = here("local/dapc_2026-01-15"), # local destination
             conflicts = "overwrite") # if copy file name already exists
```

Read `dapc_SEA`.
```{r}
dapc_SEA <- readRDS(here("local/dapc_2026-01-15/dapc_SEA_2026-01-15.rds"))
```

### Get Centroids
```{r}
centroids <- st_centroid(vn_provinces_eq)

centroid_coords <- st_coordinates(centroids)
centroid_df <- data.frame(centroid_coords)

centroid_df$province <- vn_provinces_eq$prov_eng

head(centroid_df)
```

### Variance Explained
How many PCs are needed to explain 99\% of variance?
```{r}
# get eigen values
eig <- dapc_SEA$dapc_final$eig

# calculate proportion and cumulative
prop_var <- eig / sum(eig)
cum_var  <- cumsum(prop_var)

# how many needed?
retain_PCs <- which(cum_var >= 0.99)[1] # 6
retain_PCs
```
  

### Match Data
Match PCs, centroid coordinates, and metadata.
```{r}
# metadata to keep
sea_meta <- seq_meta %>%
  mutate(year = as.integer(sub(".*-(\\d{4})$", "\\1", collection)),
         province = location) %>%
  select(year, province, host, accession)


# select first PCs and accession
K <- retain_PCs # from variance explained
pc_cols <- paste0("PC_", 1:K) # PCs to keep

sea_data_in <- dapc_SEA$PCs %>%
  select(accession, all_of(pc_cols))

# match PCs and meta
sea_data_in <- left_join(sea_data_in, sea_meta, by="accession")

# match data and coordinates
sea_data_in <- left_join(sea_data_in, centroid_df, by="province")

sea_data_in <-sea_data_in %>%
  mutate(set = "virus")

head(sea_data_in)
```

## Prepare Spatial Mesh

### Define Spatial Boundaries
```{r}
# dissolve internal polygons, keeping only the outermost country boundary
vn_provinces_union <- st_union(vn_provinces_eq)

# create convex hull around whole country
vn_hull <- st_convex_hull(vn_provinces_union)

# buffer distance
vn_buffer <- st_buffer(vn_hull,
                        dist = 15)

# convert to Spatial (sp) class (INLA required)
vn_buffer <- as(vn_buffer, "Spatial")

vn_buffer_lines <- inla.sp2segment(vn_buffer)
```


### Create Mesh
```{r fig.width=12, fig.height=15}
set.seed(1976)
mesh_dom <- inla.mesh.2d(boundary = vn_buffer_lines, 
                        loc = sea_data_in[,c("X","Y")],
                        cutoff = 10, 
                        max.edge = c(20, 50),
                        offset = c(50,120),
                        min.angle = 30) 

mesh_dom$n

# plot to check
plot(mesh_dom, main=" ")
plot(vn_provinces_eq, add=T, border = "blue", lwd=2)
plot(model_points, add=T, pch=19, size=5, col = "red")
```

### Create Template Raster
Raster version of the study area
```{r fig.width=10, fig.height=15}
tmp_ras <- raster::raster(res = 2, crs = equal_area_proj)
raster::extent(tmp_ras) <- raster::extent(vn_provinces_eq)

vn_rast <- raster::rasterize(vn_provinces_eq,
                              tmp_ras,
                              field = 0,
                              background = NA)
# check result
plot(vn_rast)
```

Dense point grid of above raster.  To be used for visualizations later.
```{r}
grid_points <- raster::rasterToPoints(vn_rast, sp = TRUE)

grid_points@data <- grid_points@data %>%
               mutate(X = grid_points@coords[,1],
                      Y = grid_points@coords[,2],
                      set = "grid") %>% # label to distinguish when combined
               select(-layer)

class(grid_points)
dim(grid_points)
head(grid_points)
```

### Estimate Virus Diversity
Within province diversity.  Higher values of `div_trace` indicate greater within province diversity. 
```{r}
prov_df <- sea_data_in %>%
  group_by(province, X, Y) %>%  # province centroids
  summarise(
    n = n(), # samples by province
    PC1_mean = mean(PC_1, na.rm = TRUE), # mean PCs in province
    PC2_mean = mean(PC_2, na.rm = TRUE),
    PC3_mean = mean(PC_3, na.rm = TRUE),
    PC4_mean = mean(PC_4, na.rm = TRUE),
    PC5_mean = mean(PC_5, na.rm = TRUE),
    PC6_mean = mean(PC_6, na.rm = TRUE),
    # within-province
    # div_trace = province diversity
    div_trace = { # covariance across PCs
      Z <- as.matrix(dplyr::pick(dplyr::all_of(pc_cols)))
      Z <- Z[complete.cases(Z), , drop = FALSE]
      if (nrow(Z) >= 2) sum(diag(stats::cov(Z))) else NA_real_
    },
    year_med = stats::median(year, na.rm = TRUE), # median year/dominant host
    host_mode = {
      h <- host[!is.na(host)]
      if (length(h) == 0) NA_character_ else names(which.max(table(h)))
    },
    .groups = "drop"
  ) %>%
  mutate(
    host_mode = factor(host_mode)
  )

```

### Singletons
There is no diversity if the province has only one sample, so NAs are in the above.  Rather than throw out single samples (singletons), we can assume that there is proportion of the average in other in other provinces.
```{r}
global_pc_var <- sum(diag(cov(as.matrix(sea_data_in[, pc_cols]),
                              use = "complete.obs")))

prov_df <- prov_df %>%
  mutate(
    div_trace_filled = case_when(
      n >= 2 ~ div_trace, # if at least two samples
      n == 1 ~ global_pc_var/K),
      log_div_trace = log(div_trace_filled) # log scale
    ) %>%
  as.data.frame()


# One province has extremely low diversity, approximately zero
# this is the min diversity to consider
eps <- quantile(prov_df$div_trace[prov_df$div_trace > 0], 0.025, na.rm = TRUE)

prov_df <- prov_df %>%
  mutate(
    div_trace_filled = pmax(div_trace_filled, eps),
    log_div_trace = log(div_trace_filled)
  )

```

### Project to Mesh
Match point locations to mesh
```{r}
# get the coordinates
locs <- cbind(prov_df$X, prov_df$Y)

# create a matrix that cross-references the point locations with the mesh.
A.mat = inla.spde.make.A(mesh_dom, 
                          alpha = 2,
                          loc=locs)

# This can be discussed separately. 
# It's one way of "guessing" at how spatially correlated data are
coords <- as.matrix(prov_df[, c("X","Y")])
nn <- FNN::get.knn(coords, k = 2)$nn.dist[,2]  # nearest neighbor distance

r0 <- stats::median(nn)*3
s0 <- sd(prov_df$PC1_mean, na.rm = TRUE)

# this is the prior for the spatial latent field.  
# Conceptually comparable to adjusting the decay rate in a kernel density function. 
spde0 <- inla.spde2.pcmatern(
  mesh_dom, alpha = 2,
  prior.range = c(r0, 0.01),
  prior.sigma = c(s0, 0.01),
  constr = TRUE
)


# this creates a coded index for locations in the mesh
latent_field <- inla.spde.make.index("latent_field", 
                               spde0$n.spde)
# unique ID for each sample
prov_df$samp_id <- paste0("samp_", 1:dim(prov_df)[1])

fmdv.lst = list(c(latent_field, # index for mesh
                list(intercept1 = 1)), # custom intercept (see other example)
                list(year = prov_df[,"year_med"],
                     host = prov_df[,"host_mode"],
                     samp_id = prov_df[,"samp_id"],
                     province = prov_df[,"province"])) 

```

### Data Stack
```{r}
fmdv.stk <- inla.stack(data = list(Y = prov_df$log_div_trace), # response/dependent variable
                                   A = list(A.mat, 1), # matrix to align with mesh
                             effects = fmdv.lst, # variables of interest  
                                 tag = "base.0") # arbitrary name/label
```


### Formulas 
First formula included `latent_field`, which is the spatial effect.  The second formula is a nonspatial version for comparison.  Comparing spatial vs. nonspatial helps us determine if accounting for spatial structure is needed.
```{r}
pcprior1 <- list(prec = list(prior="pc.prec", # prior for iid terms
                             param = c(1, 0.01))) # flat/noninformative

fam_gauss <- list(
  hyper = list( # prior for the Gaussian distribution
    prec = list(prior = "pc.prec", param = c(1, 0.01))
  )
)

# spatial formula
formula_0 <- Y ~ -1 + intercept1 +  # response and intercept  
                f(latent_field, # spatial field
                  model=spde0) +
                f(year, # sample year
                   model="iid", 
                   constr=TRUE,
                   hyper=pcprior1) +
                f(host, # hosts
                   model="iid", 
                   constr=TRUE,
                   hyper=pcprior1) +
                f(province, # province
                   model="iid", 
                   constr=TRUE,
                   hyper=pcprior1)

# non-spatial formula
formula_1 <- Y ~ -1 + intercept1 +  # response and intercept  
                #f(latent_field, # spatial field - turned OFF
                #  model=spde0) +
                f(year, # sample year
                   model="iid", 
                   constr=TRUE,
                   hyper=pcprior1) +
                f(host, # hosts
                   model="iid", 
                   constr=TRUE,
                   hyper=pcprior1) +
                f(province, # province
                   model="iid", 
                   constr=TRUE,
                   hyper=pcprior1)
```

### Run Models
```{r}
# spatial
model.spatial <- inla(formula_0, 
                     data = inla.stack.data(fmdv.stk), 
                     family = "gaussian", 
                     control.family = fam_gauss,
                     verbose = FALSE,
                     control.fixed = list(prec = 1, prec.intercept=1), 
                     control.predictor = list(
                                            A = inla.stack.A(fmdv.stk), 
                                            compute = TRUE, 
                                            link = 1), 
                     control.inla = list(strategy="adaptive", 
                                         int.strategy = "eb"),
                     control.compute=list(dic = TRUE, cpo = TRUE,
                                          waic = TRUE, config=TRUE)) 

# non spatial
model.nonspatial <- inla(formula_1, 
                         data = inla.stack.data(fmdv.stk), 
                         family = "gaussian", 
                         control.family = fam_gauss,
                         verbose = FALSE,
                         control.fixed = list(prec = 1, prec.intercept=1), 
                         control.predictor = list(
                                                A = inla.stack.A(fmdv.stk), 
                                                compute = TRUE, 
                                                link = 1), 
                         control.inla = list(strategy="adaptive", 
                                             int.strategy = "eb"),
                         control.compute=list(dic = TRUE, cpo = TRUE, 
                                              waic = TRUE, config=TRUE))

```


### Compare Models
Comparing WAICs and leave-one-out predictive scoring (LPML). If the spatial model performs better, than virus diversity is spatially structured.   
   
The LPML is a type of leave one out, predictive scoring.  Higher LPML implies better out-of-sample predictive performance. 
```{r}
# WAICs 
model.spatial$waic$waic
model.nonspatial$waic$waic

# LPML
lpml_spatial <- sum(log(model.spatial$cpo$cpo))
lpml_nonspatial <- sum(log(model.nonspatial$cpo$cpo))

c(lpml_spatial = lpml_spatial,
  lpml_nonspatial = lpml_nonspatial,
  delta_lpml = lpml_spatial - lpml_nonspatial)
```

### Unsampled Locations
```{r}
Apred <- inla.spde.make.A(mesh_dom, # map point grid to study area
                          loc = cbind(grid_points[,"X"], 
                                      grid_points[,"Y"]))

# get estimate for latent spatial field
grid_points$srf_m <- drop(Apred %*% model.spatial$summary.random$latent_field$mean) 

#get intercept
beta0 <- model.spatial$summary.fixed["intercept1", "mean"]

grid_points$eta_mean <- (beta0 + grid_points$srf_m) # this is log scale


# assign values in points to raster grid by location, i.e., "rasterize"
div_r <- rasterize(grid_points, # points with results
                   vn_rast, # grid to match to
                   "eta_mean", # which variable in the points to use
                   background = NA) # value for cells surrounding study area

```

### Province Estimates
Extracting grid values to province polygons
```{r}
div_r <- rast(div_r) # change format for extract

# extract values for polygons
stats <- extract(
    div_r,
    vn_provinces_eq,
    fun = c("min","mean","max"),
    touches = TRUE,
    na.rm = TRUE
  )

# add to province polygons
vn_provinces_eq <- cbind(vn_provinces_eq, stats)
```


## Plot Spatial Estimates  

### Grid Version
```{r fig.width=10, fig.height=15}
rast_df <- as.data.frame(div_r, xy = TRUE, na.rm = TRUE)

ggplot() +
  geom_raster(data = rast_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_viridis_c(option = "turbo", name = "Diversity") +
  geom_sf(data = vn_provinces_eq, fill = NA, color = "black", size = 2) + 
  coord_sf() +
  labs(
    title = "Estimated Diversity",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm"),
    legend.position = "right",
    legend.key.width = unit(1, "line"),
    legend.key.height = unit(2, "line"),
    legend.text = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 22, face = "bold")
  )
```

### Polygon Version
```{r fig.width=10, fig.height=15}
ggplot(vn_provinces_eq) +
  geom_sf(aes(fill = mean_layer), color = "grey30", linewidth = 0.2) +
  scale_fill_viridis_c(option = "turbo", name = "Diversity") +
  coord_sf(crs = st_crs(vn_provinces_eq)) +
  labs(
    title = "Estimated Diversity",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm"),
    legend.position = "right",
    legend.key.width = unit(1, "line"),
    legend.key.height = unit(2, "line"),
    legend.text = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 22, face = "bold")
  )
```


## PC Interpolation
The above provides an estimate of diversity.  Below, individual PCs are mapped.

### Project to Mesh
Match point locations to mesh
```{r}
# get the coordinates
locs <- cbind(prov_df$X, prov_df$Y)

# create a matrix that cross-references the point locations with the mesh.
A.mat = inla.spde.make.A(mesh_dom, 
                          alpha = 2,
                          loc=locs)

# this creates a coded index for locations in the mesh
latent_field <- inla.spde.make.index("latent_field", 
                               spde0$n.spde)
# unique ID for each sample
prov_df$samp_id <- paste0("samp_", 1:dim(prov_df)[1])

fmdv.lst = list(c(latent_field, # index for mesh
                list(intercept1 = 1)), # custom intercept (see other example)
                list(year = prov_df[,"year_med"],
                     host = prov_df[,"host_mode"],
                     samp_id = prov_df[,"samp_id"],
                     province = prov_df[,"province"])) 
```

### PC Loop
```{r}
formula_pc <- Y ~ -1 + intercept1 +  # response and intercept  
                f(latent_field, # spatial field
                  model=spde0) +
                f(year, # sample year
                   model="iid", 
                   constr=TRUE,
                   hyper=pcprior1)  

run_PCs <- paste0("PC", 1:retain_PCs, "_mean")

pc_stack <- stack()

for(i in 1:length(run_PCs)){
  
  target_pc <- prov_df[[run_PCs[i]]]
  
  loop.stk <- inla.stack(data = list(Y = target_pc), # response
                                   A = list(A.mat, 1), # matrix to align with mesh
                             effects = fmdv.lst, # variables of interest  
                                 tag = "base.0") # arbitrary name/label
  
  pc_model <- inla(formula_pc, 
                     data = inla.stack.data(loop.stk), 
                     family = "gaussian", 
                     control.family = fam_gauss,
                     verbose = FALSE,
                     control.fixed = list(prec = 1, prec.intercept=1), 
                     control.predictor = list(
                                            A = inla.stack.A(loop.stk), 
                                            compute = TRUE, 
                                            link = 1), 
                     control.inla = list(strategy="adaptive", 
                                         int.strategy = "eb"),
                     control.compute=list(dic = TRUE, cpo = TRUE,
                                          waic = TRUE, config=TRUE)) 
  
  # map result
  grid_points$srf_m <- drop(Apred %*% pc_model$summary.random$latent_field$mean) 

  #get intercept
  beta0 <- pc_model$summary.fixed["intercept1", "mean"]
  
  grid_points$eta_mean <- (beta0 + grid_points$srf_m) # this is log scale
  
  
  # assign values in points to raster grid by location, i.e., "rasterize"
  PC_r <- rasterize(grid_points, # points with results
                     vn_rast, # grid to match to
                     "eta_mean", # which variable in the points to use
                     background = NA) # value for cells surrounding study area
  
  # stack results
  pc_stack <- stack(pc_stack, PC_r)
}
  

names(pc_stack) <- run_PCs
```

### Extract to Province
```{r}
pc_stack <- rast(pc_stack) # change format for extract

# extract values for polygons
stats <- extract(
    pc_stack,
    vn_provinces_eq,
    fun = c("mean"),
    touches = TRUE,
    na.rm = TRUE
  )

# add to province polygons
vn_provinces_eq <- cbind(vn_provinces_eq, stats)
```


### Write Results
```{r}
diversity_SEA <- vn_provinces_eq %>%
  mutate(geometry = st_centroid(geometry),
         div_min = min_layer,
         div_mean = mean_layer,
         div_max = max_layer) %>%
  bind_cols(as.data.frame(st_coordinates(.))) %>%
  st_drop_geometry() %>%
  select(-c(min_layer, mean_layer,max_layer))

# save copy to repo
write.csv(diversity_SEA, here("data/diversity_SEA.csv"), row.names=FALSE) 
```

### Save Grids
```{r}
names(div_r) <- "diversity"
all_grids <- c(pc_stack, div_r)

writeRaster(
  all_grids,
  filename = here("local/SEA_grids.tif"),
  overwrite = TRUE
)

# push to OSF
osf_upload(osf_project_data, 
           conflicts = "overwrite",
           path = here("local/SEA_grids.tif"))
```


