---
title: "Disaggregation of DAPC Outputs"
format:
  gfm:
   toc: true
   toc-depth: 2
   code-fold: show
   code-summary: "Hide code"
   preview-mode: raw
editor: source
editor_options: 
  chunk_output_type: console
---

## Overview  
This script builds on the [phylogenetic regression](https://github.com/geoepi/fmd-vietnam/blob/main/example-trees/phylo_regression.md) and [spatial modeling](https://github.com/geoepi/fmd-vietnam/blob/main/example-trees/spatial_intro.md), and [DAPC](https://github.com/geoepi/fmd-vietnam/blob/main/example-trees/dapc_intro.md) examples examples.  Its objective is to *disaggregate* or subsample numeric values specific to individual, discrete, and relatively larger geographic locations to smaller geographic units. In this particular instance, the idea is to take values representing FMDv variation within Vietnam provinces and estimate a continuously varying coverage as the resolution of 25x25~km$^2$ grid cells.  Ideally, the resulting grids will show the relationships between and among different provinces. 
  
## Libraries  
Load needed packages.  
```{r message=FALSE, warning=FALSE}
library(here) # directory management
library(tidyverse) # general wrangling
library(ape) # phylodata manipulation
library(phangorn) # phylodata manipulation
library(msa) # phylodata manipulation
library(adegenet) # DAPC package 
library(sp) # spatial data manipulation
library(spdep) # spatial data manipulation
library(terra) # spatial data manipulation
library(sf) # spatial data manipulation
library(tidyverse) # general wrangling
library(viridis) # color pallete
library(ggtree) # plot trees
library(patchwork) # combine plots
library(cowplot) # pull legend from ggplot
library(geodata) # import spatial covaraites
library(INLA) # Model inference
library(Matrix) # diagnostics to check matrices
library(ggforce) # help with pie chart plots
library(ggpubr) # help with pie chart plots
library(raster)
```

## Load Sequences
Loading the VP1 sequences produced from the [VP1 trimming example](https://github.com/geoepi/fmd-vietnam/blob/main/example-trees/identify_VP1_region.md).
```{r}
alignment <- read.dna(here("example-trees/example_data/vp1_sequences_trimmed.fasta"), 
                      format = "fasta")

alignment
```

## Numeric Conversion  
Convert to a binary SNP matrix. The binary matrix is a numerical version where each position (site) in the alignment represents a Single Nucleotide Polymorphism (SNP).  
  
Make a copy of the alignment.    
```{r}
snp_matrix <- alignment
```

This chunk is mostly just reorganizing the data to identify departures from the *reference* condition, where the reference is defined as the most frequent allele at each site.  If an allele matches the reference it's coded as a 0, and if it's different it's coded as a 1.  
   
Instead of using the *most frequent* approach, the code could be tweaked to use a specific reference sequence.  There are positives and negatives either way.  We can discuss these.  
```{r}
# identify polymorphic sites
polymorphic_sites <- ape::seg.sites(snp_matrix)

# subset the alignment to include only polymorphic sites
snp_matrix_poly <- snp_matrix[, polymorphic_sites]

# convert to a binary SNP matrix
binary_snp_matrix <- apply(as.character(snp_matrix_poly), 2, function(site) {
    # find the most common nucleotide - the "reference" allele
    ref_allele <- names(sort(table(site), decreasing = TRUE))[1]
    as.numeric(site != ref_allele)  # code a 0 if it matches the reference, otherwise 1
})

# ensure proper matrix structure (sometimes things get scrambled)
binary_snp_matrix <- matrix(binary_snp_matrix, 
                            ncol = length(polymorphic_sites), 
                            byrow = FALSE)

# label row names and col names
rownames(binary_snp_matrix) <- rownames(snp_matrix)
colnames(binary_snp_matrix) <- paste0("SNP_", seq_len(ncol(binary_snp_matrix)))

# check output
dim(binary_snp_matrix)  # rows should be same as sequence number in alignment
head(rownames(binary_snp_matrix))  # row names are accession numbers
head(colnames(binary_snp_matrix))  # columns are the SNPs

binary_snp_matrix[1:10,1:6] # look at a few entries
```


## Province Metadata  
Creating a data frame (`location_lookup`) to store the metadata for the sequences used in this analysis.  
```{r}
load(here("data/prep-for-seqs.Rda"))
head(seq_meta)

location_lookup <- as.data.frame(rownames(binary_snp_matrix)) # accession #'s
names(location_lookup) <- "accession"

location_lookup$order <- 1:nrow(location_lookup) # in case the data gets shifted

location_lookup <- left_join(location_lookup, seq_meta, by = "accession")
sum(is.na(location_lookup$location)) # any missing locations? Should be 0

head(location_lookup)
```


## Supervised DAPC
Running the DAPC with the [**adegenet** package](https://github.com/thibautjombart/adegenet).  Here, the province names are used to define groups, i.e., *supervised* clustering.      
     
**NOTE:** The [DAPC](https://github.com/geoepi/fmd-vietnam/blob/main/example-trees/dapc_intro.md) example looks at DAPC results in more detail than shown here.  Skipping these important steps here to reduce clutter and better show the overall workflow.  
```{r}
set.seed(1976)

dapc_result.1 <- dapc(binary_snp_matrix, # input genomic info
                      grp = location_lookup$location, # the groups (provinces)  
                      n.pca = 20, # how many PCs to use 
                      n.da = 5) # how many DFs to return
```

Check initial results:  
Proportion of sequences that can be correctly assigned to the groups given in the *grp* input. This means that some sequences are more similar to provinces other than the one where it was collected.
```{r}
summary(dapc_result.1)$assign.prop 
```

Step 2: Optimize the number of PCs.
```{r}
optim_pcs <- optim.a.score(dapc_result.1) # find the optimal number of PCs
optim_pcs$best # here's how many suggested to avoid overfitting

# run again, different n.pca
dapc_result.2 <- dapc(binary_snp_matrix, 
                    grp= location_lookup$location,
                    n.pca = optim_pcs$best, n.da = 5)

summary(dapc_result.2)$assign.prop # about the same
```



## Spatial Analysis  
Results from the DAPC can be used for modeling.  The code below is intended to demonstrate how DAPC results can be extracted from the DAPC object, visualized, and included in downstream models. 

## Vietnam Geographic Boundaries  
Merging smaller jurisdictional boundaries into large province-level units.
```{r}
load(here("data/vn-spat.Rda"))
class(vn_spat1)


# aggregate to province
vn_provinces <- vn_spat1 %>%
  group_by(prov_eng) %>%  
  summarize(geometry = st_union(geometry), .groups = "drop") 
```

Project the long/lat coordinates to equal-areas measured in square kilometers. 
```{r}
# UTM
utm_zone_48n_km <- "+proj=utm +zone=48 +datum=WGS84 +units=km +no_defs"

# transform boundaries
vn_provinces <- st_transform(vn_provinces, crs = utm_zone_48n_km)
```

Quick visual check
```{r fig.height=8, fig.width=8}
ggplot(vn_provinces) +
  geom_sf(aes(fill=prov_eng)) +
  scale_fill_viridis_d(name="Province") +              
  theme_minimal() +
  theme(legend.position = "right")
```

### Extract Coordinates
```{r}
centroids <- st_centroid(vn_provinces)
centroid_coords <- st_coordinates(centroids)
centroid_df <- data.frame(centroid_coords)
colnames(centroid_df) <- c("long", "lat")

centroid_df$location <- vn_provinces$prov_eng

head(centroid_df)
```


### Load Province Metadata
```{r}
load(here("data/prep-for-seqs.Rda"))

head(seq_meta)
```



## Extract DAPC Results
The PCs can be extracted from the DAPC object like this:
```{r}
unsup_pcs <- dapc_result.2$tab
ncol(unsup_pcs) # number of optimal PCs retained
nrow(unsup_pcs) # number of sequences/samples

# rename to remove special characters
names(unsup_pcs) <- paste0("PC_", 1:ncol(unsup_pcs))

# add accession number to data
unsup_pcs$accession <- rownames(unsup_pcs)

# add predicted provinces
unsup_pcs$pred_prov <- dapc_result.2$assign

head(unsup_pcs)
```

Add province and other metadata to PCs
```{r}
unsup_pcs <- left_join(unsup_pcs, seq_meta, by = "accession")
```

Might also add the grographic centroid coordinates.
```{r}
unsup_pcs <- left_join(unsup_pcs, centroid_df, by = "location")

head(unsup_pcs)
```

The DFs are recorded as LDs (linear discriminant) and can be extracted from the DAPC object like this:
```{r}
unsup_dfs <- as.data.frame(dapc_result.2$ind.coord)
ncol(unsup_dfs) # number of DFs retained
nrow(unsup_dfs) # number of sequences/samples

# add accession number to data
unsup_dfs$accession <- rownames(unsup_dfs)

head(unsup_dfs)
```

Add the predicted group assignments.  These are in the same order as other data in the DAPC object.
```{r}
unsup_dfs$group <- paste0("Group_", dapc_result.2$assign)

head(unsup_dfs)
```

### Combine all data
```{r}
all_dapc_df <- left_join(unsup_dfs, unsup_pcs, by="accession")

head(all_dapc_df)
```


## Mapping

### Map DAPC identfied Groups  
This plot of discrete group assignments is only to demonstrate that there are many mismatches between the the actual province where the sample was collected and the one estimated or assigned by the DAPC.  Although the estimates are not very good, note that there is still some grouping by location.  That is, there appears to be a geographic pattern.
```{r fig.width=10, fig.height=10}
group_proportions <- all_dapc_df %>%
  group_by(location, pred_prov) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(location) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

pie_chart_data <- left_join(group_proportions, centroid_df, by = "location")

pie_chart_data <- pie_chart_data %>%
  group_by(location) %>%
  mutate(proportion = proportion / sum(proportion)) %>%
  ungroup()

pie_chart_data <- pie_chart_data %>%
  group_by(location) %>%
  mutate(
    start = cumsum(lag(proportion, default = 0)) * 2 * pi,
    end = cumsum(proportion) * 2 * pi
  ) %>%
  ungroup()


ggplot() +
  geom_sf(data = vn_provinces, fill = "gray95", color = "black", size = 0.5) +
  geom_arc_bar(
    data = pie_chart_data,
    aes(
      x0 = long, y0 = lat, r0 = 0, r = 30, 
      start = start, end = end, fill = pred_prov
    ),
    inherit.aes = FALSE
  ) +
  scale_fill_viridis_d(option = "turbo", name = "Supervised Groups") +  
  coord_sf() +
  labs(
    title = " ",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold")
  )

```

### Map LD Values  
Just as the discrete group outputs, which are nominal variables (province names), can be mapped, so too can the numeric continuous values encoded by the PCs and LDs. 
   
Ignoring the issue of multiple different samples and results from each province for a moment, first find means:
```{r}
mean_ld_values <- all_dapc_df %>%
  group_by(location) %>%
  summarise(
    mean_LD1 = mean(LD1, na.rm = TRUE),
    mean_LD2 = mean(LD2, na.rm = TRUE),
    mean_LD3 = mean(LD3, na.rm = TRUE),
    mean_LD4 = mean(LD4, na.rm = TRUE),
    mean_LD5 = mean(LD5, na.rm = TRUE)
  )

which(is.na(mean_ld_values))
```
   
then merge with the spatial polygons:
```{r}
vn_provinces_merged <- vn_provinces %>%
  left_join(mean_ld_values, by = c("prov_eng" = "location")) %>%
  mutate(
    fill_color = case_when(
      !is.na(mean_LD1) ~ "valid",  # locations with data
      TRUE ~ "missing"              # no samples/values
    )
  )
```
   
lastly, replicate the geometry for a multipanel plot:  
```{r}
geometry <- st_geometry(vn_provinces_merged)  # extract the geometry
vn_provinces_no_geom <- vn_provinces_merged %>% select(-geometry)  # remove geometry

# to long format
vn_provinces_long <- vn_provinces_no_geom %>%
  pivot_longer(cols = starts_with("mean_LD"),  # columns starting with "mean_LD"
               names_to = "LD",                # a new column
               values_to = "mean_value")       # label as mean_value

# replicate the geometry for each LD
geometry_rep <- rep(geometry, each = 5)  # 3 variables (mean_LD1, mean_LD2, mean_LD3)

# long data with the geometry column
vn_provinces_merged_long <- cbind(vn_provinces_long, geometry = geometry_rep)

# verify it's an sf object
vn_provinces_merged_long <- st_as_sf(vn_provinces_merged_long)

# verify geometries are valid
st_is_valid(vn_provinces_merged_long)  # Should return TRUE for all rows


```

and plot:
```{r fig.width=12, fig.height=10}
ggplot(data = vn_provinces_merged_long, 
       aes(fill = mean_value), 
       color = "black", size = 0.5) +
  geom_sf() +
  scale_fill_viridis_c(
    name = "Mean LD",
    option = "turbo", 
    na.value = "white" # provinces w/no samples
  ) +
  coord_sf() +
  labs(
    title = "LD by Location",
    x = "Longitude",
    y = "Latitude"
  ) +
  facet_wrap(~LD, ncol=5) +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm"),
    legend.position = "right",
    legend.key.width = unit(1, "line"),
    legend.key.height = unit(2, "line"),
    legend.text = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 22, face = "bold")
  )
```


# Spatial Model
 
### Expand data  
Ensure provinces without samples are represented in the data frame. 
```{r}
length(unique(all_dapc_df$location)) # provinces in DAPC
length(unique(centroid_df$location)) # provinces in polygons

model_data <- all_dapc_df %>%
  select(-c(long, lat)) 

model_data <- left_join(centroid_df, model_data, by = "location") %>%
  mutate(set = "centers")

length(unique(model_data$location)) # provinces #
head(model_data) # coords for all, but NA for unsampled locations
```

## Spatial Points
Although the **sf** package is the new standard for spatial analysis in R, some packages like **r-INLA** still depend on older packages. This chunk converts the model_data to a spatial points data frame.
```{r warning=FALSE, eval=TRUE}

model_points <- sp::SpatialPointsDataFrame(model_data[,c("long","lat")], model_data)
proj4string(model_points) <- utm_zone_48n_km
class(model_points)

# ol'skool spdf polygons
vn_provinces_sp <- as(vn_provinces, "Spatial")
class(vn_provinces_sp)

# check center points
plot(vn_provinces_sp, col="gray80")
plot(model_points, col="red", pch = 1, add=TRUE)

```

## Study Bounds  
A bit tedious with the **sf** package, but needed due to the [r-spatial Evolution](https://r-spatial.org/).
```{r}
# convert to sf class
vn_provinces_sf <- st_as_sf(vn_provinces_sp)

# dissolve internal polygons, keeping only the outermost country boundary
vn_provinces_union <- st_union(vn_provinces_sf)

# create convex hull around whole country
vn_hull <- st_convex_hull(vn_provinces_union)

# buffer distance
vn_buffer <- st_buffer(vn_hull,
                        dist = 15)

# convert back to Spatial (sp) class
vn_buffer <- as(vn_buffer, "Spatial")
```


## Mesh
```{r warning=FALSE, eval=TRUE, fig.height=10, fig.width=10}
vn_buffer_lines <- inla.sp2segment(vn_buffer)

set.seed(1976)
mesh_dom <- inla.mesh.2d(boundary = vn_buffer_lines, 
                        loc = model_points,
                        cutoff = 10, 
                        max.edge = c(20, 50),
                        offset = c(50,120),
                        min.angle = 30) 

mesh_dom$n
plot(mesh_dom, main=" ")
plot(vn_provinces_sp, add=T, border = "blue", lwd=2)
plot(model_points, add=T, pch=19, size=5, col = "red")
```

## Vietnam Raster
```{r}
tmp_ras <- raster::raster(res = 2, crs = utm_zone_48n_km)
raster::extent(tmp_ras) <- raster::extent(vn_provinces_sp)

vn_rast <- raster::rasterize(vn_provinces_sp,
                              tmp_ras,
                              field = 0,
                              background = NA)

# random values to view
vn_rast_rand <- vn_rast

values(vn_rast_rand)[!is.na(values(vn_rast_rand))] <- rnorm(sum(!is.na(values(vn_rast_rand))), 0, 2)

# quick check
#plot(vn_rast_rand)
#plot(vn_provinces_sp, add=T, border = "blue", lwd=2)
```

### View Raster
```{r fig.height=10, fig.width=10}
rast_df <- as.data.frame(vn_rast_rand, xy = TRUE, na.rm = TRUE)

vn_provinces_sf <- st_as_sf(vn_provinces_sp)

ggplot() +
  geom_raster(data = rast_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_viridis_c(option = "turbo", name = "Random Values") +  
  geom_sf(data = vn_provinces_sf, fill = NA, color = "black", size = 2) + 
  coord_sf() +
  labs(
    title = "Vietnam Raster",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm"),
    legend.position = "right",
    legend.key.width = unit(1, "line"),
    legend.key.height = unit(2, "line"),
    legend.text = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 22, face = "bold")
  )
```
  
### Extract raster coordinates
```{r}
grid_points <- raster::rasterToPoints(vn_rast, sp = TRUE)

grid_points@data <- grid_points@data %>%
               mutate(long = grid_points@coords[,1],
                      lat = grid_points@coords[,2],
                      set = "grid") %>% # label to distinguish when combined
               dplyr::select(-layer)


class(grid_points)
head(grid_points@data)
```

### Combine Samples, Centroids, and Grid Cells 
One big data set with raster cells and samples represented.  Unsampled locations are coded with NA due to no data being available.
```{r}
missing_cols <- setdiff(names(model_data), names(grid_points@data))
grid_points@data[missing_cols] <- NA

model_data <- rbind(grid_points@data, model_data)

dim(model_data)

head(model_data)
tail(model_data)

# convert back to spatial points
model_points <- sp::SpatialPointsDataFrame(model_data[,c("long","lat")], model_data)
proj4string(model_points) <- utm_zone_48n_km
class(model_points)
```

## Covariates
Using the same raster grid variables from the [spatial modeling](https://github.com/geoepi/fmd-vietnam/blob/main/example-trees/spatial_intro.md) example, which includes more detail and explanation.  
  
The independent or explanatory variables are in a gridded format. The grids are first transformed to match the study area and geographic projection. Then, the code extracts the grid values at point locations. 
  
### Buffalo Density
```{r}
# Buffalo Density
buffalo_density <- rast(here("local/livestock/GLW4-2020.D-DA.BFL.tif"))
buffalo_density

buffalo_density <- project(buffalo_density, proj4string(vn_provinces_sp))
buffalo_density <- crop(buffalo_density, vn_provinces_sp)

model_points_vect <- vect(model_points@data, geom = c("long", "lat"))

# quick check, ensure overlap
# plot(buffalo_density)
# plot(model_points, add=TRUE, col="red")

model_points_vect$buff_density <- extract(buffalo_density, model_points_vect, 
                                          method= "bilinear", na.rm=T)[,"GLW4-2020.D-DA.BFL"]
```

### Minimum temperature  
Same basic steps for minimum temper
```{r}
tmin <- geodata::worldclim_country("Vietnam", var="tmin",
                                   path = here("local/"))

vn_tmin <- min(tmin) # calculate min across all months, i.e., average annual minimum
vn_tmin

vn_tmin <- project(vn_tmin, proj4string(vn_provinces_sp))
vn_tmin <- crop(vn_tmin, vn_provinces_sp)

# quick check, ensure overlap
# plot(vn_tmin)
# plot(model_points, add=TRUE, col="red")

model_points_vect$tmin <- extract(vn_tmin, model_points_vect, method= "bilinear", na.rm=T)[,"min"]
```

A few grid cell points near the shoreline fall in the ocean.  Those are removed here:
```{r}
valid_points <- which(!is.na(model_points_vect$tmin) & !is.nan(model_points_vect$tmin) &
                      !is.na(model_points_vect$buff_density) & !is.nan(model_points_vect$buff_density))

model_points_vect <- model_points_vect[valid_points, ]
```

For ease of model fitting, the two grid variables are scaled and centered.  
```{r}
model_points_vect$tmin_sc <- as.numeric(
  scale(model_points_vect$tmin, center = T, scale = T)
)

model_points_vect$buff_density_sc <- as.numeric(
  scale(model_points_vect$buff_density, center = T, scale = T)
)
```


## Mesh Projection
This code links the point coordinate locations to the correct and corresponding position in the triangulated mesh constructed earlier.  
```{r}
# get data from spatial object
model_points_df <- as.data.frame(model_points_vect)
vect_coords <- geom(model_points_vect)[,c("x","y")]
colnames(vect_coords) <- c("long", "lat")

# combine
model_points_df <- cbind(vect_coords, model_points_df)

# Check the resulting data frame
head(model_points_df)

# get the coordinates
locs <- cbind(model_points_df$long, model_points_df$lat)

# create a matrix that cross-references the point locations with the mesh.
A.mat = inla.spde.make.A(mesh_dom, 
                          alpha = 2,
                          loc=locs)

# This can be discussed separately. One way of "guessing" at how spatially correlated data are
range_lat <- diff(range(model_points_df$lat))/3

# this is the prior for the spatial latent field.  
# Conceptually comparable to adjusting the decay rate in a kernel density function. 
spde0 <- inla.spde2.pcmatern(mesh_dom, alpha = 2,
                            prior.range=c(range_lat, 0.01),  
                            prior.sigma=c(1, 0.01),
                            constr = TRUE)

# this creates a coded index for locations in the mesh
latent_field <- inla.spde.make.index("latent_field", 
                               spde0$n.spde)
```

## Data Stack  
The r-INLA package works best if data is organized as data as a list() object:
```{r}
model_points_df$ID <- 1:nrow(model_points_df) # unique id for every location

fmdv.lst = list(c(latent_field, # index for mesh
                list(intercept1 = 1)), # custom intercept (see other example)
                list(tmin = model_points_df[,"tmin_sc"], # variables of interest
                     buff = model_points_df[,"buff_density_sc"],
                     ID = model_points_df[,"ID"],
                     set = model_points_df[,"set"],
                     host = model_points_df[,"host"],
                     location = model_points_df[,"location"])) 

```


### Stack Object  
This list holds the other list, as well as the response variable, mesh projection, matrix, and a name or label. 
```{r}
fmdv.stk <- inla.stack(data = list(Y = model_points_df$LD3), # response/dependent variable
                                   A = list(A.mat, 1), # matrix to align data with mesh locations
                             effects = fmdv.lst, # variables of interest  
                                 tag = "base.0") # arbitrary name/label to pull results later
```


## Formula
```{r}
pcprior1 <- list(prec = list(prior="pc.prec", param = c(1, 0.01))) 

formula_0 <- Y ~ -1 + intercept1 +  # response and intercept  
				f(latent_field, # spatial field
          model=spde0) +
				f(ID, # individual point variation
				   model="iid", 
				   constr=TRUE,
				   hyper=pcprior1) +
       f(host, # hosts
				   model="iid", 
				   constr=TRUE,
				   hyper=pcprior1) +
				tmin + buff # variables from grid
```


## Run Model  
Model run may take 2-5 minutes.  
```{r warning=FALSE, message=FALSE, eval=FALSE}
model.0 <- inla(formula_0, 
				 data = inla.stack.data(fmdv.stk), 
				 family = "gaussian", 
				 verbose = FALSE,
				 control.fixed = list(prec = 1, prec.intercept=1), 
				 control.predictor = list(
										A = inla.stack.A(fmdv.stk), 
										compute = TRUE, 
										link = 1), 
				 control.inla = list(strategy="adaptive", 
									 int.strategy = "eb"),
				 control.compute=list(dic = TRUE, cpo = TRUE, waic = TRUE)) 

# save, if needed
# saveRDS(model.0, here("local/models_out/model0.rds"))
```
```{r}
# loadfrom prior run
model.0 <- readRDS(here("local/models_out/model0.rds"))
  
summary(model.0)
```

## Host Effect  
View results of the random effect for host species as given in sample metadata.
```{r fig.height=10, fig.width=10}
host_effect <- model.0$summary.random$host[,1:6]
colnames(host_effect) <- c("Host", "mean", "sd", "Quant0.025","Quant0.5", "Quant0.975" )

ggplot(host_effect, aes(Host, Quant0.5)) +
  geom_hline(yintercept = 0,
             col="red",
             linewidth= 1,
             linetype = "dotdash") +
  geom_point(col="black") +
  geom_linerange(aes(ymin = Quant0.025, ymax = Quant0.975), col="black") +
  labs(
      title = " ",
      x = " ",
      y = "Host Effect",
      fill = " "
    ) +
  coord_flip() +
  theme_minimal() +
    theme(
      plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm"),
      legend.position = "none",
      legend.key.size = unit(2, "line"),
      legend.text = element_text(size = 16, face = "bold"),
      legend.title = element_text(size = 18, face = "bold"),
      axis.title.x = element_text(size = 22, face = "bold"),
      axis.title.y = element_text(size = 22, face = "bold"),
      axis.text.x =  element_text(size = 18, face = "bold"),
      axis.text.y = element_text(size = 10, face = "bold"),
      plot.title = element_text(size = 22, face = "bold")
    )
```
  
### Get the fitted values  
These are the estimates for all point locations in the input data.
```{r}
# create an index to filter results to only the input data
data_index = inla.stack.index(fmdv.stk, "base.0")$data

# add the results to the points data object (so it can be overlaid on the raster)
model_points_vect$fit_LD3 <- model.0$summary.fitted.values$mean[data_index]

# convert format - AGAIN!
rasterize_set <- as(model_points_vect, "Spatial")

# can filter to only grid points or use everything
rasterize_set <- subset(rasterize_set, set == "grid")

# assign values in points to raster grid by location, i.e., "rasterize"
LD3_r <- rasterize(rasterize_set, # points with results
                   vn_rast, # grid to match to
                   fun = "max", # what to do if more than one point is in a raster cell
                   "fit_LD3", # which variable in the points to use
                   background = NA) # value for cells surrounding study area

# save raster, if desired
# writeRaster(LD3_r , here("local/grids_out/LD3_est.tif"), overwrite = TRUE)
# xx <- raster::raster(here("local/grids_out/LD3_est.tif"))
```


## Map Result  
Continuous version of response variable as estimated from input explanatory variables and the spatial pattern of samples.  Locations without samples were listed as NA in the response variable so made no contribution to the estimate.  However, variables like tmin and buffalo density did because those values were know for all locations.  
```{r fig.width=10, fig.height=10}
rast_df <- as.data.frame(LD3_r, xy = TRUE, na.rm = TRUE)

ggplot() +
  geom_raster(data = rast_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_viridis_c(option = "turbo", name = "LD3",
                       limits = c(-1, 1.5)) +  
  geom_sf(data = vn_provinces_sf, fill = NA, color = "black", size = 2) + 
  coord_sf() +
  labs(
    title = "Estimated LD",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm"),
    legend.position = "right",
    legend.key.width = unit(1, "line"),
    legend.key.height = unit(2, "line"),
    legend.text = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 22, face = "bold"),
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 22, face = "bold")
  )

save(rast_df, vn_provinces_sf, file=here("example-trees/example_data/dapc-plot.Rda"))
```


#### The above workflow can be mofified for other LDs or PCs and to include other explanatory variables.
