---
title: "sequence-map tutorial workbook"
subtitle: "From viral sequence alignment to spatial genetic-variation surfaces"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
execute:
  warning: true
  message: true
---

## Interpretation first

This workbook maps **ordination-axis scores**, beginning with PCA. A PCA axis
surface is a spatial prediction of genetic-position scores, not a direct
estimate of nucleotide diversity. Location-level reduced-space dispersion
requires at least two sequences at a sampled location. DAPC is optional because
it depends on grouping assumptions; inferred groups should be treated as
exploratory. Predictions away from sampled locations are model-based
interpolations, not observed genetic data.

## Setup

```{r}
#| label: setup
#| include: false

library(sequencemap)
library(dplyr)
library(ggplot2)
library(sf)

repo_root <- normalizePath(
  if (file.exists("DESCRIPTION")) getwd() else file.path(getwd(), ".."),
  winslash = "/",
  mustWork = TRUE
)
fasta_path <- file.path(repo_root, "data", "fasta", "vp1_A_trimmed.fasta")
metadata_path <- file.path(repo_root, "data", "metadata", "prep-for-seqs.Rda")
boundary_path <- file.path(repo_root, "data", "spatial", "vn-spat.Rda")
output_dir <- file.path(repo_root, "outputs", "workbook-sequence-map")

sequence_id_col <- "sequence_id"
lon_col <- "longitude"
lat_col <- "latitude"
location_col <- "location"
crs_input <- 4326
crs_projected <- "+proj=aea +lat_1=20 +lat_2=40 +lat_0=30 +lon_0=105 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
grid_resolution <- 25000
n_pca <- 20
run_dapc <- FALSE
write_outputs <- FALSE
inla_available <- requireNamespace("INLA", quietly = TRUE)

materialize_repository_examples <- function(metadata_rda, boundary_rda) {
  metadata_env <- new.env(parent = emptyenv())
  boundary_env <- new.env(parent = emptyenv())
  load(metadata_rda, envir = metadata_env)
  load(boundary_rda, envir = boundary_env)

  metadata <- get("seq_meta_reg", envir = metadata_env)
  provinces <- get("vn_provinces", envir = boundary_env)
  province_points <- sf::st_transform(
    sf::st_point_on_surface(sf::st_geometry(sf::st_transform(provinces, 3857))),
    4326
  )
  coordinates <- sf::st_coordinates(province_points)
  location_coordinates <- data.frame(
    location = provinces$prov_eng,
    longitude = coordinates[, "X"],
    latitude = coordinates[, "Y"],
    stringsAsFactors = FALSE
  )

  match_index <- match(metadata$location, location_coordinates$location)
  metadata$longitude <- location_coordinates$longitude[match_index]
  metadata$latitude <- location_coordinates$latitude[match_index]
  names(metadata)[names(metadata) == "accession"] <- "sequence_id"

  metadata_file <- tempfile(fileext = ".csv")
  boundary_file <- tempfile(fileext = ".gpkg")
  utils::write.csv(metadata, metadata_file, row.names = FALSE)
  sf::st_write(provinces, boundary_file, quiet = TRUE, delete_dsn = TRUE)
  list(metadata_file = metadata_file, boundary_file = boundary_file)
}

example_files <- materialize_repository_examples(metadata_path, boundary_path)
```

The bundled repository data use province-level representative coordinates for
this instructional example. They are adequate for demonstrating the workflow,
but are not a substitute for original sampling coordinates.

## Read and validate inputs

```{r}
alignment <- read_alignment(fasta_path)
metadata <- read_metadata(example_files$metadata_file)
boundary <- read_boundary(example_files$boundary_file)

character_alignment <- ape::as.character.DNAbin(alignment)
if (is.list(character_alignment)) {
  character_alignment <- do.call(rbind, character_alignment)
}

alignment_dimensions <- dim(character_alignment)
alignment_ids <- names(alignment)
represented_ids <- sum(alignment_ids %in% metadata[[sequence_id_col]])

validate_alignment_metadata_match(alignment, metadata, id_col = sequence_id_col)
validate_sequence_metadata(
  metadata = metadata,
  alignment = alignment,
  id_col = sequence_id_col,
  longitude_col = lon_col,
  latitude_col = lat_col
)

data.frame(
  alignment_sequences = alignment_dimensions[1],
  alignment_sites = alignment_dimensions[2],
  metadata_rows = nrow(metadata),
  metadata_columns = ncol(metadata),
  alignment_ids_represented = represented_ids
)
```

```{r}
cat("Boundary CRS:", sf::st_crs(boundary)$input, "\n")
print(sf::st_bbox(boundary))
```

## Alignment encoding diagnostics

```{r}
variant_result <- alignment_to_variant_matrix(alignment, drop_invariant = TRUE)
variant_matrix <- variant_result$variant_matrix
site_summary <- variant_result$site_summary
sequence_summary <- variant_result$sequence_summary

data.frame(
  n_sequences = nrow(variant_matrix),
  n_original_sites = nrow(site_summary),
  n_retained_sites = sum(site_summary$retained),
  n_encoded_columns = ncol(variant_matrix)
)
```

```{r}
site_class_counts <- site_summary %>%
  count(site_class, name = "n_sites") %>%
  arrange(desc(n_sites))
site_class_counts
```

```{r}
ggplot(site_class_counts, aes(x = site_class, y = n_sites)) +
  geom_col(fill = "#2C7FB8") +
  labs(x = NULL, y = "Number of alignment sites", title = "Alignment site classes") +
  theme_minimal()
```

```{r}
qc_columns <- intersect(c("n_gap", "n_ambiguous", "n_missing"), names(sequence_summary))
if (length(qc_columns) > 0L) {
  sequence_summary %>%
    select(sequence_id, all_of(qc_columns)) %>%
    tidyr::pivot_longer(-sequence_id, names_to = "state", values_to = "count") %>%
    ggplot(aes(x = count)) +
    geom_histogram(bins = 20, fill = "#7FCDBB", color = "white") +
    facet_wrap(~state, scales = "free_y") +
    labs(x = "Count per sequence", y = "Number of sequences", title = "Sequence QC states") +
    theme_minimal()
}
```

## Principal component analysis

```{r}
pca_result <- run_sequence_pca(variant_matrix, rank. = n_pca)
pca_scores <- pca_result$scores
pca_result$variance_summary %>% slice_head(n = 10)
```

```{r}
ggplot(pca_result$variance_summary, aes(x = component, y = proportion_variance, group = 1)) +
  geom_col(fill = "#41AB5D") +
  geom_line(color = "#238B45") +
  labs(x = "Principal component", y = "Proportion variance", title = "PCA scree plot") +
  theme_minimal()
```

```{r}
ggplot(pca_result$variance_summary, aes(x = component, y = cumulative_variance, group = 1)) +
  geom_line(color = "#756BB1") +
  geom_point(color = "#756BB1") +
  labs(x = "Principal component", y = "Cumulative variance", title = "Cumulative PCA variance") +
  theme_minimal()
```

```{r}
pca_plot_data <- pca_scores %>%
  left_join(metadata %>% select(all_of(c(sequence_id_col, location_col))), by = c("sequence_id" = sequence_id_col))

ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = .data[[location_col]])) +
  geom_point(size = 2, alpha = 0.8) +
  labs(color = "Location", title = "PC1 and PC2 sequence scores") +
  theme_minimal()
```

## Optional discriminant analysis of principal components

DAPC is disabled by default. Use it only with biologically meaningful group
labels, or regard inferred clusters as exploratory.

```{r}
#| eval: false

run_dapc <- FALSE
group_col <- "region"
if (run_dapc) {
  dapc_result <- run_sequence_dapc(
    x = variant_matrix,
    metadata = metadata,
    id_col = sequence_id_col,
    grouping_col = group_col,
    infer_groups = FALSE,
    n_pca = n_pca,
    n_da = 3
  )
}
```

## Location summaries

```{r}
sequence_scores <- pca_scores %>% select(sequence_id, matches("^PC_?[0-9]+$"))
location_scores <- summarize_location_scores(
  scores = sequence_scores,
  metadata = metadata,
  sequence_id_col = sequence_id_col,
  location_col = location_col,
  lon_col = lon_col,
  lat_col = lat_col
)
location_diversity <- summarize_location_diversity(
  scores = sequence_scores,
  metadata = metadata,
  sequence_id_col = sequence_id_col,
  location_col = location_col,
  lon_col = lon_col,
  lat_col = lat_col
)

data.frame(n_sampled_locations = nrow(location_scores))
```

```{r}
ggplot(location_scores, aes(x = n_sequences)) +
  geom_histogram(binwidth = 1, boundary = 0, fill = "#FEC44F", color = "white") +
  labs(x = "Sequences per location", y = "Number of locations") +
  theme_minimal()
```

```{r}
location_diversity %>%
  filter(n_sequences >= 2) %>%
  select(location, n_sequences, n_axes, div_trace, mean_pairwise_ordination_distance)
```

```{r}
ggplot(filter(location_diversity, n_sequences >= 2), aes(x = reorder(location, div_trace), y = div_trace)) +
  geom_col(fill = "#E34A33") +
  coord_flip() +
  labs(x = NULL, y = "Ordination-space dispersion (trace)") +
  theme_minimal()
```

## Spatial preparation

All mesh and prediction operations use a projected CRS in metres. The Vietnam
Albers CRS selected above is a local analysis projection; choose a projection
appropriate to the geographic extent of a new study.

```{r}
points_sf <- prepare_spatial_points(
  data = location_scores,
  lon_col = lon_col,
  lat_col = lat_col,
  crs_input = crs_input,
  crs_projected = crs_projected,
  id_col = location_col,
  keep_cols = setdiff(names(location_scores), c(lon_col, lat_col))
)
prediction_grid <- make_prediction_grid(
  boundary = boundary,
  crs_projected = crs_projected,
  grid_resolution = grid_resolution,
  return = "sf"
)
boundary_projected <- sf::st_transform(boundary, sf::st_crs(points_sf))
```

```{r}
ggplot() +
  geom_sf(data = boundary_projected, fill = "grey95") +
  geom_sf(data = points_sf, color = "#D7301F", size = 2) +
  labs(title = "Sampled locations") +
  theme_minimal()
```

```{r}
ggplot() +
  geom_sf(data = boundary_projected, fill = "grey95") +
  geom_sf(data = prediction_grid, color = "#2B8CBE", size = 0.3, alpha = 0.5) +
  geom_sf(data = points_sf, color = "#D7301F", size = 1.5) +
  labs(title = "Prediction grid and sampled locations") +
  theme_minimal()
```

## INLA availability

```{r}
if (!requireNamespace("INLA", quietly = TRUE)) {
  knitr::asis_output("INLA is not installed; mesh and model sections are skipped.")
} else {
  knitr::asis_output("INLA is available; mesh, model, and prediction sections are evaluated.")
}
```

## Mesh diagnostics

```{r}
#| eval: !expr inla_available

mesh_result <- make_inla_mesh(
  points_sf = points_sf,
  boundary = boundary,
  prediction_grid = prediction_grid,
  max_edge = c(grid_resolution * 2, grid_resolution * 5),
  cutoff = grid_resolution / 5,
  offset = c(grid_resolution, grid_resolution * 2),
  crs_projected = crs_projected
)
mesh_result$mesh_summary
```

With only 17 sampled locations in this example, mesh density should be read as
a numerical approximation choice, not as evidence for fine-scale genetic
structure.

```{r}
#| eval: !expr inla_available

plot(mesh_result$mesh, asp = 1, main = "Full INLA mesh")
plot(sf::st_geometry(boundary_projected), add = TRUE, border = "grey40")
points(sf::st_coordinates(points_sf), col = "#D7301F", pch = 16)
```

```{r}
#| eval: !expr inla_available

point_coordinates <- sf::st_coordinates(points_sf)
plot(mesh_result$mesh, asp = 1, main = "Mesh near sampled locations", xlim = range(point_coordinates[, 1]) + c(-50000, 50000), ylim = range(point_coordinates[, 2]) + c(-50000, 50000))
points(point_coordinates, col = "#D7301F", pch = 16)
```

## Axis models

```{r}
#| eval: !expr inla_available

response_cols <- intersect(c("PC1_mean", "PC2_mean", "PC3_mean"), names(points_sf))
axis_models <- fit_axis_models(
  points_sf = points_sf,
  response_cols = response_cols,
  mesh = mesh_result,
  continue_on_error = TRUE
)
axis_models$model_summaries
axis_models$successful_response_cols
axis_models$failed_response_cols
```

WAIC and DIC are model diagnostics for a given axis. They do not validate a
biological interpretation of that axis or establish population-genetic change.

## Spatial predictions

```{r}
#| eval: !expr inla_available

axis_surfaces <- predict_axis_surfaces(
  axis_models = axis_models,
  prediction_grid = prediction_grid,
  continue_on_error = TRUE
)
names(axis_surfaces$combined_surface)
```

```{r}
#| eval: !expr inla_available

plot_prediction <- function(surface, column, title) {
  ggplot() +
    geom_sf(data = surface, aes(color = .data[[column]]), size = 0.6) +
    geom_sf(data = boundary_projected, fill = NA, color = "grey30") +
    geom_sf(data = points_sf, color = "black", size = 1) +
    labs(color = column, title = title) +
    theme_minimal()
}

for (column in intersect(c("PC1_mean_pred", "PC2_mean_pred", "PC3_mean_pred"), names(axis_surfaces$combined_surface))) {
  print(plot_prediction(axis_surfaces$combined_surface, column, paste("Predicted", column)))
}
```

```{r}
#| eval: !expr inla_available

if ("PC1_mean_sd" %in% names(axis_surfaces$combined_surface)) {
  plot_prediction(axis_surfaces$combined_surface, "PC1_mean_sd", "PC1 prediction uncertainty")
}
```

## Raster outputs and polygon summaries

Polygon summaries are reporting summaries of continuous model predictions.
They are not separate province-level models and are not direct nucleotide
diversity estimates. Keep within-polygon predicted-axis heterogeneity, model
uncertainty, and observed sequence dispersion as distinct quantities.

```{r}
#| eval: !expr inla_available

axis_raster <- axis_surface_to_raster(axis_surfaces$combined_surface)
province_summaries <- aggregate_axis_rasters_to_polygons(
  raster = axis_raster,
  polygons = boundary,
  polygon_id_col = "prov_eng"
)

province_display_cols <- c(
  "polygon_id",
  "n_grid_cells",
  intersect(
    c(
      "PC1_mean_pred_mean", "PC1_mean_pred_sd",
      "PC2_mean_pred_mean", "PC2_mean_pred_sd",
      "PC3_mean_pred_mean", "PC3_mean_pred_sd",
      "PC1_mean_sd_mean", "PC2_mean_sd_mean", "PC3_mean_sd_mean"
    ),
    names(province_summaries)
  )
)
province_summaries %>% select(all_of(province_display_cols))
```

```{r}
#| eval: !expr inla_available

if ("PC1_mean_pred_mean" %in% names(province_summaries)) {
  province_map <- left_join(
    boundary,
    province_summaries,
    by = c("prov_eng" = "polygon_id")
  )
  ggplot(province_map) +
    geom_sf(aes(fill = PC1_mean_pred_mean)) +
    labs(fill = "Mean PC1 prediction", title = "Province reporting summary for PC1") +
    theme_minimal()
}
```

```{r}
#| eval: !expr inla_available

if (write_outputs) {
  raster_outputs <- write_axis_geotiffs(
    axis_surfaces,
    output_dir = file.path(output_dir, "rasters")
  )
  utils::write.csv(
    province_summaries,
    file.path(output_dir, "province_axis_surface_summaries.csv"),
    row.names = FALSE
  )
}
```

## Optional small outputs

```{r}
if (write_outputs) {
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
  utils::write.csv(site_summary, file.path(output_dir, "alignment_site_summary.csv"), row.names = FALSE)
  utils::write.csv(sequence_summary, file.path(output_dir, "alignment_sequence_summary.csv"), row.names = FALSE)
  utils::write.csv(location_scores, file.path(output_dir, "location_scores.csv"), row.names = FALSE)
  utils::write.csv(location_diversity, file.path(output_dir, "location_diversity.csv"), row.names = FALSE)
}
```

## Reproducibility

```{r}
packageVersion("sequencemap")
sessionInfo()
```
