---
title: "Vietnam sequence-map analysis"
subtitle: "Spatial genetic-variation surfaces across four FMDv alignments"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
execute:
  warning: true
  message: true
editor_options: 
  chunk_output_type: console
---

## Analysis purpose

This is a Vietnam analysis workbook, not the package tutorial. It
applies the PCA-first sequence-map workflow independently to every FASTA
alignment discovered in data/fasta. The current analysis expects four
alignments, but discovery is dynamic and does not assume an exact count.

Each alignment has its own ordination basis. Province-level values are derived reporting summaries of continuous SPDE surfaces, not direct nucleotide diversity and not separate province-level models.

## Settings

```{r}
#| label: setup
#| include: false

library(here)
library(sequencemap)
library(dplyr)
library(ggplot2)
library(sf)

repo_root <- here::here()

run_models <- TRUE
overwrite_existing <- TRUE
rebuild_reporting_outputs <- TRUE
n_pca <- 20
n_axes_model <- 3
grid_resolution <- 10000
crs_input <- 4326
crs_projected <- "+proj=aea +lat_1=20 +lat_2=40 +lat_0=30 +lon_0=105 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"
output_root <- here::here("outputs", "vietnam-analysis")
write_outputs <- TRUE

metadata_rda <- here::here("data", "metadata", "prep-for-seqs.Rda")
boundary_rda <- here::here("data", "spatial", "vn-spat.Rda")

required_reporting_functions <- c(
  "write_axis_geotiffs",
  "aggregate_axis_rasters_to_polygons",
  "calculate_polygon_axis_metrics"
)
missing_reporting_functions <- required_reporting_functions[
  !vapply(
    required_reporting_functions,
    exists,
    logical(1),
    where = asNamespace("sequencemap"),
    inherits = FALSE
  )
]
if (length(missing_reporting_functions) > 0L) {
  message(
    "The installed sequencemap package is missing: ",
    paste(missing_reporting_functions, collapse = ", "),
    ". Reinstall the current repository before running or rebuilding reporting outputs."
  )
}

materialize_vietnam_examples <- function(metadata_rda, boundary_rda) {
  metadata_env <- new.env(parent = emptyenv())
  boundary_env <- new.env(parent = emptyenv())
  load(metadata_rda, envir = metadata_env)
  load(boundary_rda, envir = boundary_env)

  metadata <- get("seq_meta_reg", envir = metadata_env)
  provinces <- get("vn_provinces", envir = boundary_env)
  province_points <- sf::st_transform(
    sf::st_point_on_surface(sf::st_geometry(sf::st_transform(provinces, 3857))),
    4326
  )
  coordinates <- sf::st_coordinates(province_points)
  location_coordinates <- data.frame(
    location = provinces$prov_eng,
    longitude = coordinates[, "X"],
    latitude = coordinates[, "Y"],
    stringsAsFactors = FALSE
  )
  match_index <- match(metadata$location, location_coordinates$location)
  metadata$longitude <- location_coordinates$longitude[match_index]
  metadata$latitude <- location_coordinates$latitude[match_index]
  names(metadata)[names(metadata) == "accession"] <- "sequence_id"

  metadata_path <- tempfile(fileext = ".csv")
  boundary_path <- tempfile(fileext = ".gpkg")
  utils::write.csv(metadata, metadata_path, row.names = FALSE)
  sf::st_write(provinces, boundary_path, quiet = TRUE, delete_dsn = TRUE)
  list(
    metadata = metadata,
    provinces = provinces,
    metadata_path = metadata_path,
    boundary_path = boundary_path
  )
}

example_inputs <- materialize_vietnam_examples(metadata_rda, boundary_rda)
```

Set run_models to TRUE only when ready to fit all alignment-specific INLA
models locally. A full run can take substantial time. Set
rebuild_reporting_outputs to TRUE to add rasters and polygon metrics to
existing workflow results without refitting models.

## Input discovery

```{r}
fasta_dir <- here::here("data", "fasta")
fasta_files <- list.files(
  fasta_dir,
  pattern = "\\.fasta$|\\.fa$|\\.fas$",
  full.names = TRUE,
  ignore.case = TRUE
)
alignment_names <- tools::file_path_sans_ext(basename(fasta_files))

alignment_manifest <- data.frame(
  alignment_name = alignment_names,
  fasta_path = fasta_files,
  output_dir = file.path(output_root, alignment_names),
  workflow_result_path = file.path(output_root, alignment_names, "workflow_result.rds"),
  prediction_surface_path = file.path(output_root, alignment_names, "prediction_axis_surfaces.csv"),
  province_metrics_path = file.path(output_root, alignment_names, "province_axis_derived_metrics.csv"),
  stringsAsFactors = FALSE
)

alignment_manifest %>%
  select(alignment_name, fasta_path, output_dir) %>%
  slice_head(n = 10)
```

## Metadata and boundary preparation

```{r}
metadata <- example_inputs$metadata
boundary <- example_inputs$provinces

data.frame(
  metadata_rows = nrow(metadata),
  n_provinces = nrow(boundary),
  boundary_crs = sf::st_crs(boundary)$input,
  n_fasta_files = nrow(alignment_manifest)
)
alignment_manifest %>% select(alignment_name, fasta_path)
```

## Run analysis loop

When models are disabled, this section only inventories expected outputs. When
enabled, it runs each FASTA alignment separately using PCA-first defaults and
then creates derived raster and province reporting products.

```{r}
write_alignment_reporting_outputs <- function(workflow_result, output_dir, boundary) {
  if (length(missing_reporting_functions) > 0L) {
    stop(
      "Reinstall the current sequencemap repository before creating reporting outputs.",
      call. = FALSE
    )
  }
  if (is.null(workflow_result$axis_surfaces$combined_surface)) {
    message("No combined axis surface is available for ", output_dir)
    return(invisible(NULL))
  }

  raster_outputs <- write_axis_geotiffs(
    workflow_result$axis_surfaces,
    output_dir = file.path(output_dir, "rasters")
  )
  province_summaries <- aggregate_axis_rasters_to_polygons(
    raster = raster_outputs$raster,
    polygons = boundary,
    polygon_id_col = "prov_eng"
  )
  polygon_metrics <- calculate_polygon_axis_metrics(province_summaries)
  utils::write.csv(
    province_summaries,
    file.path(output_dir, "province_axis_surface_summaries.csv"),
    row.names = FALSE
  )
  utils::write.csv(
    polygon_metrics,
    file.path(output_dir, "province_axis_derived_metrics.csv"),
    row.names = FALSE
  )
  invisible(list(
    raster_outputs = raster_outputs,
    province_summaries = province_summaries,
    polygon_metrics = polygon_metrics
  ))
}

run_status <- alignment_manifest %>%
  transmute(
    alignment_name,
    output_dir,
    status = if_else(dir.exists(output_dir), "existing output directory", "not run")
  )

if (run_models) {
  if (length(missing_reporting_functions) > 0L) {
    stop(
      "The installed sequencemap package is outdated. Reinstall this repository before setting run_models = TRUE.",
      call. = FALSE
    )
  }
  if (!requireNamespace("INLA", quietly = TRUE)) {
    stop("INLA is required when run_models = TRUE.", call. = FALSE)
  }

  for (index in seq_len(nrow(alignment_manifest))) {
    entry <- alignment_manifest[index, , drop = FALSE]
    if (dir.exists(entry$output_dir) && !overwrite_existing) {
      message("Skipping existing output directory for ", entry$alignment_name)
      run_status$status[index] <- "skipped existing output"
      next
    }

    workflow_result <- run_sequence_map_workflow(
      fasta_path = entry$fasta_path,
      metadata_path = example_inputs$metadata_path,
      boundary_path = example_inputs$boundary_path,
      output_dir = entry$output_dir,
      sequence_id_col = "sequence_id",
      lon_col = "longitude",
      lat_col = "latitude",
      location_col = "location",
      crs_input = crs_input,
      crs_projected = crs_projected,
      grid_resolution = grid_resolution,
      n_pca = n_pca,
      n_axes_model = n_axes_model,
      run_dapc = FALSE,
      overwrite = overwrite_existing
    )

    realized_output_dir <- workflow_result$output_dir
    write_alignment_reporting_outputs(workflow_result, realized_output_dir, boundary)
    saveRDS(workflow_result, file.path(realized_output_dir, "workflow_result.rds"))
    run_status$status[index] <- "completed"
  }
} else if (rebuild_reporting_outputs) {
  for (index in seq_len(nrow(alignment_manifest))) {
    entry <- alignment_manifest[index, , drop = FALSE]
    if (!file.exists(entry$workflow_result_path)) {
      run_status$status[index] <- "workflow result not found"
      next
    }
    workflow_result <- readRDS(entry$workflow_result_path)
    write_alignment_reporting_outputs(workflow_result, entry$output_dir, boundary)
    run_status$status[index] <- "reporting outputs rebuilt"
  }
} else {
  message(
    "Models are disabled. Set run_models <- TRUE to run the full analysis, ",
    "or rebuild_reporting_outputs <- TRUE to post-process existing results."
  )
}

run_status
```

## Cross-alignment summary

```{r}
load_workflow_summary <- function(entry) {
  if (!file.exists(entry$workflow_result_path)) {
    return(data.frame(
      alignment_name = entry$alignment_name,
      n_sequences = NA_integer_,
      n_sites = NA_integer_,
      n_retained_sites = NA_integer_,
      n_sampled_locations = NA_integer_,
      n_prediction_grid_cells = NA_integer_,
      successful_axes = NA_character_,
      failed_axes = NA_character_,
      output_dir = entry$output_dir,
      stringsAsFactors = FALSE
    ))
  }

  result <- readRDS(entry$workflow_result_path)
  data.frame(
    alignment_name = entry$alignment_name,
    n_sequences = result$alignment_qc$n_sequences,
    n_sites = result$alignment_qc$n_sites,
    n_retained_sites = result$alignment_qc$n_retained_sites,
    n_sampled_locations = nrow(result$location_scores),
    n_prediction_grid_cells = nrow(result$prediction_grid),
    successful_axes = paste(result$axis_models$successful_response_cols, collapse = ", "),
    failed_axes = paste(result$axis_models$failed_response_cols, collapse = ", "),
    output_dir = result$output_dir,
    stringsAsFactors = FALSE
  )
}

cross_alignment_summary <- bind_rows(lapply(
  seq_len(nrow(alignment_manifest)),
  function(index) load_workflow_summary(alignment_manifest[index, , drop = FALSE])
))
cross_alignment_summary
```

## Workflow audit status

```{r}
if (!exists(
  "summarize_workflow_status",
  where = asNamespace("sequencemap"),
  inherits = FALSE
)) {
  stop(
    "The installed sequencemap package does not include workflow audit functions. ",
    "Reinstall the current repository, restart R, and rerun this workbook.",
    call. = FALSE
  )
}

load_workflow_audit <- function(entry) {
  if (!file.exists(entry$workflow_result_path)) {
    return(NULL)
  }
  workflow_result <- readRDS(entry$workflow_result_path)
  audit <- workflow_result$workflow_status
  if (is.null(audit)) {
    audit <- summarize_workflow_status(workflow_result)
  }
  audit
}

workflow_audits <- lapply(
  seq_len(nrow(alignment_manifest)),
  function(index) load_workflow_audit(alignment_manifest[index, , drop = FALSE])
)
names(workflow_audits) <- alignment_manifest$alignment_name

for (alignment_name in names(workflow_audits)) {
  audit <- workflow_audits[[alignment_name]]
  if (is.null(audit)) {
    next
  }
  cat("### ", alignment_name, "\n\n", sep = "")
  print(
    audit$workflow_status %>%
      select(
      output_dir,
      n_sequences,
      n_retained_sites,
      n_sampled_locations,
      n_prediction_grid_cells,
      dapc_status,
      n_axis_models_successful,
      n_axis_models_failed,
      n_axis_predictions_successful,
      n_axis_predictions_failed,
      raster_status,
        polygon_status
      )
  )
  print(
    audit$axis_model_status %>%
      select(response_col, status, n_observations, n_mesh_vertices, waic, dic, error_message) %>%
      slice_head(n = 10)
  )
  print(
    audit$axis_prediction_status %>%
      select(response_col, status, n_prediction_cells, error_message) %>%
      slice_head(n = 10)
  )
  print(audit$error_log %>% slice_head(n = 10))
}
```

## Province metric comparison

```{r}
combined_province_metrics <- bind_rows(lapply(
  seq_len(nrow(alignment_manifest)),
  function(index) {
    entry <- alignment_manifest[index, , drop = FALSE]
    if (!file.exists(entry$province_metrics_path)) {
      return(NULL)
    }
    metrics <- utils::read.csv(entry$province_metrics_path, stringsAsFactors = FALSE)
    metrics$alignment_name <- entry$alignment_name
    metrics
  }
))

if (nrow(combined_province_metrics) == 0L) {
  message("No province metric files were found yet.")
} else {
  if (write_outputs) {
    dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
    utils::write.csv(
      combined_province_metrics,
      file.path(output_root, "vietnam_combined_province_axis_metrics.csv"),
      row.names = FALSE
    )
  }

  show_top_provinces <- function(metric) {
    combined_province_metrics %>%
      filter(!is.na(.data[[metric]])) %>%
      group_by(alignment_name) %>%
      slice_max(.data[[metric]], n = 10, with_ties = FALSE) %>%
      ungroup() %>%
      select(alignment_name, polygon_id, n_grid_cells, all_of(metric))
  }

  show_top_provinces("axis_centroid_distance")
  show_top_provinces("axis_surface_dispersion")
  show_top_provinces("mean_axis_uncertainty")
}
```

## Maps

```{r}
plot_alignment_outputs <- function(entry) {
  surface_path <- file.path(entry$output_dir, "prediction_axis_surfaces.gpkg")
  metrics_path <- entry$province_metrics_path

  if (file.exists(surface_path)) {
    surface <- sf::st_read(surface_path, quiet = TRUE)
    for (column in intersect(
      c("PC1_mean_pred", "PC2_mean_pred", "PC3_mean_pred"),
      names(surface)
    )) {
      print(
        ggplot() +
          geom_sf(data = surface, aes(color = .data[[column]]), size = 0.4) +
          geom_sf(data = boundary, fill = NA, color = "grey30") +
          labs(title = paste(entry$alignment_name, column), color = column) +
          theme_minimal()
      )
    }
  }

  if (file.exists(metrics_path)) {
    metrics <- utils::read.csv(metrics_path, stringsAsFactors = FALSE)
    province_map <- left_join(boundary, metrics, by = c("prov_eng" = "polygon_id"))
    for (metric in intersect(
      c("axis_centroid_distance", "axis_surface_dispersion", "mean_axis_uncertainty"),
      names(province_map)
    )) {
      province_map$plot_value <- province_map[[metric]]
      print(
        ggplot(province_map) +
          geom_sf(aes(fill = plot_value)) +
          labs(title = paste(entry$alignment_name, metric), fill = metric) +
          theme_minimal()
      )
    }
  }
}

invisible(lapply(
  seq_len(nrow(alignment_manifest)),
  function(index) plot_alignment_outputs(alignment_manifest[index, , drop = FALSE])
))
```

## Interpretation notes

PCA signs and directions are arbitrary within each alignment.   
Province summaries are derived reporting summaries. Interpret observed
location-level reduced-space dispersion separately from modeled province
metrics and posterior uncertainty. 

## Output inventory

```{r}
inventory_files <- c(
  workflow_result = "workflow_result.rds",
  axis_models = "axis_models.rds",
  axis_surfaces = "axis_surfaces.rds",
  prediction_surface = "prediction_axis_surfaces.csv",
  province_summaries = "province_axis_surface_summaries.csv",
  province_metrics = "province_axis_derived_metrics.csv",
  multiband_geotiff = file.path("rasters", "sequence_map_axis_surface_multiband.tif")
)

output_inventory <- bind_rows(lapply(
  seq_len(nrow(alignment_manifest)),
  function(index) {
    entry <- alignment_manifest[index, , drop = FALSE]
    data.frame(
      alignment_name = entry$alignment_name,
      file_type = names(inventory_files),
      path = file.path(entry$output_dir, unname(inventory_files)),
      present = file.exists(file.path(entry$output_dir, unname(inventory_files))),
      stringsAsFactors = FALSE
    )
  }
))
output_inventory
```

## Reproducibility

```{r}
packageVersion("sequencemap")
sessionInfo()
```
